Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-08-08 resubmission/paper-blinded-resubmit.tex
Up carpeta resubmit

\title{Perceptions of Inequality and Meritocracy: Their Interplay in
Shaping Preferences for Market Justice in Chile (2016-2023)}

\date{}
\begin{document}
\maketitle
\begin{abstract}
This study investigates the relationship between perceptions of economic
inequality, meritocratic perceptions, and preferences for market justice
in Chile between 2016 and 2023. Using six waves of panel data from the
Chilean Longitudinal Social Survey - ELSOC (\(N_{observations}\) =
8,643; \(N_{individuals}\) = 1,687), the analysis examines how
subjective assessments of inequality shape attitudes toward the role of
merit in access to key social services such as healthcare, education,
and pensions. Results indicate that rising perceptions of inequality are
associated with lower support for market justice; however, individuals
with strong meritocratic convictions are more likely to legitimize
existing disparities. The study also considers the influence of major
social movements during this period, which appear to have reshaped
public discourse and perceptions of fairness. These findings contribute
to a deeper understanding of how beliefs about justice and equity evolve
in contexts marked by persistent inequality and entrenched
market-oriented frameworks. \newline \textbf{Keywords}: Economic
inequality, meritocracy, market justice, Chile, public preferences,
inequality perception
\end{abstract}

@@

This study tests several key hypotheses concerning the relationships
between perceptions of economic inequality, meritocratic perceptions,
and preferences toward market justice in Chile. It posits that higher
perceived inequality and stronger meritocratic perceptions both
independently contribute to greater support for market-based
distributions of social goods. Furthermore, the study examines how
perceived economic inequality moderates the influence of meritocratic
perceptions on market justice preferences. Additionally, it considers
how major social and political mobilizations, such as those occurring
around 2019 and subsequent years, may have weakened these relationships,
reflecting increased critical perspectives on systemic inequalities and
market mechanisms. The subsequent discussion interprets the empirical
findings in light of these hypotheses, considering their implications
for understanding attitudes toward inequality and social justice in a
context marked by profound social change.

The first set of hypothesis suggested that perceptions of economic
inequality were strongly linked to preferences for market justice. The
findings demonstrated that individuals perceiving higher income
disparities showed less support for the idea that the access to core
social services depends on individual income, aligning with theories
emphasizing that awareness of inequality could fuel a critical view of
such policies (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al., 2019};
\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). Such a relationship
reflects a moral economy in which perceptions of systemic unfairness
undermine the legitimacy of existing distributions and fuel demands for
greater equity. In particular, the negative within-individual effect
observed over time (\(H1_{b}\)) suggests that as inequality becomes more
salient for some individuals, their support for market justice
declines---potentially due to rising distrust in market mechanisms or
growing disillusionment with the fairness of the system. However,
although the effect of perceived inequality at the between-individual
level is also negative, it is not statistically significant. Therefore,
we cannot conclude that individuals who, on average, perceive higher
inequality are systematically less supportive of market justice. As a
result, hypothesis \(H1_{a}\) does not receive empirical support.

Regarding the second set of hypotheses, the results confirmed that
meritocratic perceptions---particularly those emphasizing individual
effort---were associated with stronger support for market-based
distribution systems. Individuals who believed that success is primarily
achieved through personal effort were more likely to justify unequal
access to core social services based on income, interpreting such
disparities as outcomes of individual merit rather than systemic
injustice (\(H2_{a}\)). This finding aligns with previous research
showing that meritocratic narratives serve as moral justifications that
legitimize social stratification
(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These perceptions
operate symbolically to reinforce structural inequalities by reducing
support for redistributive policies, framing inequality as both fair and
deserved. Consistent with prior work by Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) on Chilean students,
the findings suggest that such meritocratic perceptions uphold existing
hierarchies by promoting the acceptance of inequality as a reflection of
individual virtue rather than structural failure. This mechanism is
particularly salient in neoliberal contexts like Chile, where market
logics heavily shape social attitudes
(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).

Interestingly, intra-individual changes in meritocratic perceptions over
time (\(H2_{b}\)) reveal mixed effects. While increases in the belief
that rewards are based on individual effort are associated with stronger
preferences for market justice, increases in the belief that rewards
derive from talent are linked to weaker support for such principles. One
possible explanation is that effort is generally viewed as a
controllable and malleable trait, aligning with the normative
foundations of market fairness. In contrast, talent tends to be
perceived as innate and less subject to personal control, rendering
talent-based inequality less legitimate. Furthermore, increased exposure
to real-world scenarios in which outcomes are clearly shaped by inherent
traits rather than hard work may lead individuals to question the
fairness and legitimacy of market-based reward systems.

The third set of hypotheses addressed the moderating role of perceptions
of inequality in the relationship between meritocratic perceptions and
preferences for market justice. The analysis revealed that the positive
association between meritocratic perceptions and support for
market-based distribution systems tends to be stronger when perceived
economic inequality is high---contradicting our initial expectations
(\(H3_{a}\) and \(H3_{b}\)). Although this moderating effect is not
statistically significant for effort-related meritocratic perceptions,
it is significant for talent-related perceptions at both the within- and
between-individual levels. This finding runs counter to our hypotheses,
as it suggests that the effect of meritocratic
perceptions---specifically, talent-based meritocracy---intensifies as
perceptions of inequality increase. However, this interpretation
requires caution. The main effect of perceived inequality on support for
market justice is negative. Thus, as perceived inequality becomes less
negative (i.e., closer to zero), the positive relationship between
talent-based meritocratic perceptions and market justice preferences
becomes stronger. This suggests a more nuanced dynamic: meritocratic
perceptions, particularly those emphasizing talent, may serve as a
stronger justificatory mechanism for market-based inequalities among
individuals who are less inclined to perceive inequality as a problem.
In other words, among those who do not strongly perceive systemic
disparities, meritocratic narratives may play a more influential role in
legitimizing unequal outcomes.

Finally, the fourth hypothesis examined the potential impact of recent
social mobilizations---particularly those that erupted in 2019---on the
relationship between meritocratic perceptions, perceptions of
inequality, and preferences for market justice. The main effect of time
indicates that, on average, support for market justice increased
following the 2019 protests. However, the analysis did not detect any
significant interactions between distributive beliefs and support for
market justice, suggesting that the social mobilizations did not
substantially modify the underlying associations between these
variables. As a result, hypotheses \(H4_{a}\) and \(H4_{b}\) are not
supported by the data. This may point to a certain stability---or even
resilience---in the normative frameworks that guide individuals'
evaluations of distributive justice, despite the occurrence of major
collective political events. Alternatively, while the protests may have
eroded trust in institutional arrangements or governance, they may not
have fundamentally altered individuals' beliefs about how rewards and
resources should be allocated in society.

>>>>

\title{Perceptions of Inequality and Meritocracy: Their Interplay in
Shaping Preferences for Market Justice in Chile (2016-2023)}
\author{ }

\date{}
\begin{document}
\maketitle
\begin{abstract}
Several countries have experienced a shift toward the privatization and
commodification of public goods, welfare policies, and social services.
In Latin America, Chile stands out as a paradigmatic case where this
trend has led to the extensive marketization of essential services. From
a moral economy perspective, the extent to which individuals consider it
fair for access to such services to depend on market criteria has been
conceptualized as market justice preferences. This study investigates
the relationship between perceptions of economic inequality,
meritocratic beliefs, and market justice preferences in Chile between
2016 and 2023. Using six waves of panel data from the Chilean
Longitudinal Social Survey -- ELSOC (\(N_{observations}\) = 8,643;
\(N_{individuals}\) = 1,687), the analysis examines how subjective
assessments of inequality shape attitudes toward the role of merit in
access to key social services such as healthcare, education, and
pensions. The findings show that greater perceived inequality is
associated with lower market justice preferences. However, individuals
who believe that effort is rewarded are more likely to legitimize
existing disparities. In contrast, the perception that talent is
rewarded shows a negative effect on market justice preferences; an
effect that intensifies as perceived inequality increases over time. The
study also considers the influence of major social movements during this
period, which appear to have reshaped public discourse on justice and
fairness. These findings contribute to a deeper understanding of how
distributive beliefs evolve in contexts marked by persistent inequality
and entrenched neoliberal frameworks \newline \textbf{Keywords}:
Inequality perception, meritocracy, commodification, Chile, public
preferences
\end{abstract}

@@

The first set of hypotheses proposed that perceptions of economic
inequality are strongly associated with preferences for market justice.
In the between-individual specification, the effect of perceived
inequality is not statistically significant. Therefore, we cannot
conclude that individuals who, on average, perceive higher levels of
inequality are systematically less supportive of market justice. As a
result, hypothesis \(H1_{a}\) is not supported by the data. However, in
the within-individual specification, the findings reveal a significant
negative association: individuals who, at a given point in time,
perceive higher income disparities are less supportive of the idea that
access to core social services should depend on individual income. This
result aligns with theories suggesting that increased awareness of
inequality fosters a more critical stance toward market-based
distributive arrangements
(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al., 2019};
\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). Such a pattern reflects a
moral economy logic in which perceptions of systemic unfairness
undermine the legitimacy of existing distributions and strengthen
demands for greater equity. Specifically, the negative within-individual
effect over time (\(H1_{b}\)) suggests that as inequality becomes more
salient for some individuals, their support for market justice
declines---possibly due to growing distrust in market mechanisms or
increasing disillusionment with the perceived fairness of the system.

Regarding the second set of hypotheses, the results confirmed that
meritocratic perceptions---particularly those emphasizing individual
effort---were associated with stronger support for market-based
distribution systems. Individuals who believed that success is primarily
achieved through personal effort were more likely to justify unequal
access to core social services based on income, interpreting such
disparities as outcomes of individual merit rather than systemic
injustice (\(H2_{a}\)). This finding aligns with previous research
showing that meritocratic narratives serve as moral justifications that
legitimize social stratification
(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These perceptions
operate symbolically to reinforce structural inequalities by reducing
support for redistributive policies, framing inequality as both fair and
deserved. Consistent with prior work by Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) on Chilean students,
the results suggest that such meritocratic perceptions uphold existing
hierarchies by promoting the acceptance of inequality as a reflection of
individual virtue rather than structural failure. This mechanism is
particularly salient in neoliberal contexts like Chile, where market
logics heavily shape social attitudes
(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).

Interestingly, intra-individual changes in meritocratic perceptions over
time (\(H2_{b}\)) reveal mixed effects. While increases in the belief
that rewards are based on individual effort are associated with stronger
preferences for market justice, increases in the belief that rewards
derive from talent are linked to weaker support for such principles. One
possible explanation is that effort is generally viewed as a
controllable and malleable trait, whereas talent tends to be perceived
as innate and less subject to personal control, rendering talent-based
inequality less legitimate. Furthermore, increased exposure to
real-world scenarios in which outcomes are clearly shaped by inherent
traits rather than hard work may lead individuals to question the
fairness and legitimacy of market-based reward systems.

The third set of hypotheses addressed the moderating role of perceptions
of inequality in the relationship between meritocratic perceptions and
preferences for market justice. The analysis revealed that the positive
association between meritocratic perceptions and support for
market-based distribution systems tends to be stronger when perceived
economic inequality is high, contradicting our initial expectations
(\(H3_{a}\) and \(H3_{b}\)). Although this moderating effect is not
statistically significant for effort-related meritocratic perceptions,
it is significant for talent-related perceptions at both the within- and
between-individual levels. However, this interpretation requires
caution: the main effect of perceived inequality on support for market
justice is negative, meaning that as perceived inequality becomes less
negative (i.e., closer to zero), the positive relationship between
talent-based meritocratic perceptions and market justice preferences
becomes stronger. This suggests a more nuanced dynamic: meritocratic
perceptions, particularly those emphasizing talent, may serve as a
stronger justificatory mechanism for market-based inequalities among
individuals who are less inclined to perceive inequality as a problem.
In other words, among those who do not strongly perceive systemic
disparities, meritocratic narratives may play a more influential role in
legitimizing unequal outcomes.

Finally, the fourth exploratory hypothesis examined the potential impact
of social events --- particularly those that erupted in 2019 --- on the
relationship between meritocratic perceptions, perceptions of
inequality, and preferences for market justice. The main effect of time
indicates that, on average, support for market justice increased after
2019 protests. However, the analysis did not detect any significant
interactions between distributive beliefs and support for market
justice, suggesting that events in this period (as the mobilizations,
the pandemic and/or the constitucional processes) did not substantially
modify the underlying associations between these variables. As a result,
hypotheses \(H4_{a}\) and \(H4_{b}\) are not supported by the analysis.
This may point to a certain stability or even resilience in the
normative frameworks that guide individuals' evaluations of distributive
justice, despite the occurrence of major collective political events.
Alternatively, while the protests may have eroded trust in institutional
arrangements or governance, they may not have fundamentally altered
individuals' beliefs about how rewards and resources should be allocated
in society.
